Modeling workflow: Lake N2O survey data

### 2023-03-17 scripts/NLA_N2O_models.Rmd
Edits to math formatting and text

---
title: "Modeling workflow: Lake $N_2O$ survey data"
author: "Roy Martin, Jake Beaulieu, Michael McManus"
date: "`r Sys.Date()`"
output:
  github_document:
    number_sections: TRUE
    df_print: "tibble"
    math_method: 
      engine: webtex
      url: https://latex.codecogs.com/svg.image?
    html_preview: TRUE
    keep_html: TRUE
bibliography: RWM_Endnote_Library.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
---

@@

# Background and Objectives
This document details the modeling workflow implemented for estimating dissolved and equilibrium N2O gas concentrations and saturation ratios for a targeted population of US lakes, which included all freshwater lakes and reservoirs in the lower 48 US states larger than 4 ha. Data for these estimates came from the 2017 Nation Lakes Assessment (NLA) survey, wherein waterbodies were sampled according to a stratified, unequal probability design. Sampling was random within categories of lake size (surface area in hectares), WSA9 ecoregion, and US state (excluding AK and HI). In addition, larger lakes were intentionally over-sampled in proportion to smaller lakes. In order to make inferences from this sample to the population of interest, the estimates needed to be adjusted according to these factors.

The "Complete Data Likelihood" is a useful conceptualization for describing how biases arise in sampling [@Link_Barker_2010; @Zachmann_etal_2022]. For the 2017 NLA survey, the complete data, or population of interest, was all US lakes and reservoirs in the lower 48 states larger than 4 hectares. In this context, the survey samples were considered a subset of the complete data. Due to the pattern of selection assigned by the survey, some lakes of certain categories (e.g., small lakes) in the target population were missing from the sample not at random, but conditional on the unequal probability design. This type of non-random "missingness" is not ignorable when making inferences from a sample to a population of interest, as it can lead to biased estimates [@Gelman_etal_2014, Ch. 8].

A relatively simple, model-based strategy to account for this type of selection bias is to include the survey design variables as predictors in a regression model and then poststratify the estimates based on a known target population distribution; an adjustment procedure often referred to as multilevel regression and poststratification [MRP, @Park_etal_2004; @Gelman_Little_1997; @Gelman_etal_2020, Ch. 17]. With a stratified design, such as the one used for the 2017 NLA, the selection scheme results in pre-specified groupings within which samples may be drawn at random. Within a particular grouping or stratification, the selected observations would be considered a simple random sample (SRS) from that group's population. A regression model conditioning the responses on that grouping structure would then result in group-specific estimates that are unbiased with respect to the groups. To make make unbiased inferences to a broader population of interest, those regression estimates may be re-weighted by the known distribution of units from those same groups in a broader target population. That distribution may be summarized in a poststratification table, which, for the 2017 NLA survey, could be a cross-tabulation of lakes in the target population along the design variables: ecoregion, state, and size category. Thus, a regression model conditioning on this design would provide estimates for groups of lakes described by a size classification within a particular state and ecoregion. To adjust the regression estimates to an estimate for the population of interest (e.g., all US lakes larger than 4 ha), those group-specific regression estimates from the sample would be re-weighted based on the proportional distribution of lakes in those groups in the target population. For a recent applied example and tutorial, see @Kennedy_Gelman_2021. For an example of a recent application in the context of national surveys of environmental resources, see @Zachmann_etal_2022.

To make population-level estimates from the 2017 NLA dissolved gas data, a model-based approach, similar to those referenced above, was followed. The specific workflow, data, models, and code used to make these estimates are documented in the remainder of this document. First, Bayesian multilevel regression models were used to fit the sample data to the design variable structure. Multilevel models are often recommended in this type of survey estimation because they provide regularized estimates along the design groupings, which can improve out-of-sample inferences [e.g., @Gelman_Little_1997; @Kennedy_Gelman_2021]. Inferences for group levels that may be missing from the sample, but are part of the population of interest are also straightforward using this approach [@Gelman_etal_2020 Ch. 17; @McElreath_2020]. The second step is typically a simple poststratification of the sample-based estimates over a table of target population weights. However, the estimates ultimately needed to be useful for estimating the total gas flux at the population level, which required lake level estimates. That is, flux estimates were dependent on lake-level measures of surface area. Therefore, instead of predicting to a postratification table, predictions were made to each individual lake in the population of interest. This meant predicting to 465,897 individual natural and man made US lakes larger than 4 hectares in the lower 48 states. These predictions were assumed relevant to average conditions during the biological index period for each lake in 2017.

The specific objectives of the modeling effort were to provide population level estimates for (1) dissolved and equilibrium N2O concentrations; (2) the N2O saturation ratio (i.e., dissolved N2O/equilibrium N2O); and (3) the proportion of under-saturated water bodies (i.e., saturation ratio < 1). The gas estimates would also be subsequently used to estimate the total flux of N2O gas attributable to the target population of lakes over the index period. The saturation ratio estimates were modeled as a derived quantity from the joint model of dissolved and equilibrium N2O and were calculated as the ratio of modeled dissolved to equilibrium N2O. Because dissolved and equilibrium N2O were observed on the same sample units (lake sites), models for estimating their joint distribution were used. That is, the response variable was multivariate to account for any potential statistical dependencies between dissolved and equilibrium N2O due to, for example, common dependencies on geography. Although point predictions of the mean marginal probabilities from separate models could provide comparable estimates, a joint model allowing correlated errors was expected to better capture uncertainty and potentially improve out-of-sample predictions, should the variables be conditionally correlated [@Warton_etal_2015; @Poggiato_etal_2021]. All of the models fit were constructed using the `brms` package [@Burkner_2017] in `R` [@R_Core_Team_2021] as an interface to Stan, a software package for fitting fully Bayesian models via Hamiltonian Monte Carlo [HMC, @Stan_Development_Team_2018_a; @Stan_Development_Team_2018_b; @Stan_Development_Team_2018_c]. More specific details on these models is provided in the "Model fitting" section below.

# Data
As explained in a previous data munging document document
(https://github.com/USEPA/DissolvedGasNla/blob/master/scripts/dgIndicatorAnalysis.html), duplicate dissolved gas samples were collected at a depth of ~0.1m at designated index sites distributed across 1091 lakes nationwide, of which 95 were sampled twice as repeat visits. This randomly selected subset of revisit sites was used as a test set for assessing model fit and out-of-sample performance.

Gas samples were analyzed via gas chromotography and concentrations were recorded to the nearest 0.001 nmol/L. The samples were collected under a stratified, unequal probability design and each gas observation was indexed to an individual lake selected with unequal probability from 5 different lake size categories, $j \in j=1,...,J = 5$, according to surface area (ha), and from within a state, $k \in k=1,...,K = 48$, situated within an aggregated, WSA9 or Omernik ecoregion, $l \in l=1,...,L = 9$. All 9 WSA9 ecoregions were represented in the sample, including Xeric (XER), Western Mountain (WMT),  Northern Plains (NPL), Southern Plains (SPL), Temperate Plains (TPL), Coastal Plains (CPL), Upper Midwest (UMW), Northern Appalachian (NAP), and Southern Appalachian (SAP) regions. As shown below, the data from the initial and revisit samples were separately compiled into data frame objects in $\textbf{R}$, with $n=984$ and $n=95$ rows, respectively, of gas observations indexed to the survey design variables and several potentially relevant covariates.

## Import
The gas data and covariates were previously described and munged at 
https://github.com/USEPA/DissolvedGasNla/blob/master/scripts/dataMunge.html. That dataset was imported below. 
```{r import_data, eval=FALSE, include=TRUE}
load( file = paste0( localPath,
              "/Environmental Protection Agency (EPA)/",
              "ORD NLA17 Dissolved Gas - Documents/",
              "inputData/dg.2021-02-01.RData")
      )

@@

From the imported dataset, a new data frame for modeling was constructed from the original file including only the variables of interest: (1) the N2O gas observations; (2) the survey design variables indexed to those observations; and (3) additional covariates considered potentially useful for improving the fit of the model. The data frame below excluded the second-visit observations, which would later be used for model checking. Some variables from the imported data were renamed for convenience. In addition, the NO3 covariate was rounded according to the documented measurement precision. An alternative version of the NO3 covariate was also created in this step by log-transforming and re-coding it as an ordered factor with five levels at hand-drawn cut points. The left-most cut point separated observations below the detection limit from the completely observed samples. The remaining cut points in the positive direction were drawn at approximately equal distances along the log scale. Finally, it should be noted that one lake that was sampled was missing information on the N2O gas measurements and it was removed from the data frame.
```{r model_data, echo=TRUE, paged.print=TRUE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/dg.rda")

@@

A second dataframe, including only the second visit observations, was constructed below. These data were later used as a "test set" to assess the out-of-sample fit of the model developed on the first-visit or training data.
```{r test_data, echo=TRUE, paged.print=TRUE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/dg.rda")

@@

## Target population
Below. the NLA sampling frame was imported and then filtered to include only the target population or sampling frame for this project.
```{r import_sample_frame, echo=TRUE}
df_pop <- read.csv(file = paste0(localPath,
              "/Environmental Protection Agency (EPA)/",
              "ORD NLA17 Dissolved Gas - Documents/",
              "inputData/NLA_Sample_Frame.csv"), header = T)

@@

The resulting target population above included a total of 465,897 waterbodies.

Cross tabulations below describe the structure of the target population with respect to the design variables. The cross-tabulation makes it clear that each ecoregion does not contain each state. Therefore, in the statistical sense, states were nested in ecoregions.
```{r frame_dimensions_1, echo=TRUE, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sframe.rda")

@@

Likewise, lake size category was nested in state (which was nested in ecoregion). That is, not every ecoregion:state in the population of interest contained every size category (below).
```{r frame_dimensions_4, echo=TRUE, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sframe.rda")

@@

Below, the sampling frame was selected down to create a post-stratification table. Some of the variables were renamed to match the naming conventions used in the observational data above. There were 536 types of lakes in the population of interest with respect to the sampling design. The counts of those lake types (n_lakes) and their proportions relative to the total population of lakes in the sampling frame (prop_cell) are indicated below.
```{r filter_frame, echo=TRUE, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sframe.rda")

@@

## Sample vs. population
Below, the lake distributions in the population of interest were compared to the proportions in the observed sample. There were 352 lake types in the sample compared to the 536 in the population of of interest. In total, there were 984 observations distributed across these 352 lake types in the sample; and the number of samples was not distributed evenly across the types. Some cells were represented by as few as 1 lake. In total, 536-352 = 184 lake types in the population of interest were not represented in the sample.
```{r sample_cell_counts, echo=TRUE, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

Below, a graphical comparison was constructed to depict the distribution of cells in the population of interest _versus_ those in the sample.
```{r compare_sample_pop_cells, echo=FALSE, fig.align='center', fig.height=4, fig.width=10}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/pframe.rda")
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/samp_props.rda")

@@

Another comparison between population and sample was constructed below by ecoregion. The samples were not balanced across ecoregions. Lakes in the Coastal Plains (CPL) ecoregion, for example, were clearly undersampled relative to their proportion of the population.
```{r eco_props_pop, eval=FALSE, include=TRUE, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/pframe.rda")

@@

A similar comparison by state was constructed below.
```{r state_props_pop, eval=FALSE, include=TRUE, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/pframe.rda")

@@

Finally, a comparison by lake size category is shown below. Note that small lakes were under-sampled relative to larger lakes by design.
```{r size_props_pop, eval=FALSE, include=TRUE, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/pframe.rda")

@@

## Sample-based estimates
The overall mean and standard deviation for N2O in the sample:
```{r sample_summary_n2o}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

The same summary for equilibrium N2O:
```{r sample_summary_n2oeq}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

The saturation ratio (i.e., N2O / N2O-eq):
```{r sample_summary_n2osat}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

Finally, roughly 67% of lakes in the sample were undersaturated (i.e., saturation ratio < 1):
```{r sample_summary_propsat}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

Using only the sample observations, a plot was constructed below of the overall mean (dashed line) along with the ecoregion-specific means (black circles). The shaded areas indicate +/- 1 standard deviation. Neither dissolved N2O nor the saturation ratio were clearly structured by ecoregion in the sample, but there did appear to be some structure in the equilibrium N2O observations.
```{r sample_summary_eco, echo=FALSE, fig.align='center', fig.height=6, fig.width=8, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

The same summary by state is below.
```{r sample_summary_state, echo=FALSE, fig.align='center', fig.height=6, fig.width=8, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

Finally, the same summary by size category:
```{r sample_summary_size, echo=FALSE, fig.align='center', fig.height=6, fig.width=8, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

## Sample data exploration
Below, the empirical distribution of N2O observations in the sample was summarized using a density and rug plot below. Note the natural log scale of the x axis. Both the N2O and equilibrium N2O data had considerable right skew even after the log transformation, which was not unexpected and has been noted in other studies [@Webb_etal_2019]. The saturation ratio was also skewed since it was derived from the other two observed variables (i.e., sat_ratio = n2o / n2o_eq).
```{r summary_N2O, echo=FALSE, fig.align='center', fig.height=8, fig.width=5, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

Below are plots of N2O vs. NO3. The first plot shows log(N2O) vs. log(NO3), as well as the ordinal categories assigned to NO3 (vertical lines). The leftmost vertical line is dashed and separates the NO3 observations below the detection limit.
```{r summary_N2O_vs_NO3, echo=FALSE, fig.align='center', fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

In the plot above, the trend is increasing and nonlinear on the log scale. The increasing variance in N2O along the NO3 gradeient suggested a potential mediator of the relationship between NO3 on N2O. Below are plots of N2O vs. NO3 for 6 quantiles of the surface temperature measurements (quantiles increasing from 1 to 6). This plot below suggested that the NO3 effect on N2O may have been stronger in lakes with higher observed temperatures.
```{r summary_N2O_vs_NO3_surftemp, echo=FALSE, fig.align='center', fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

The next plot below shows the relationship between N2O and NO3 at 6 different quantiles (increasing 1 to 6) of the log-scaled lake surface area estimates.
```{r summary_N2O_vs_NO3_logarea, echo=FALSE, fig.align='center', fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

wherein $\varrho$ is the group-level residual correlation between responses.

>>>>

---
title: "Modeling workflow: Lake N2O survey data"
author: "Roy Martin, Jake Beaulieu, Michael McManus"
date: "`r Sys.Date()`"
output:
  github_document:
    number_sections: TRUE
    df_print: "tibble"
    math_method: 
      engine: webtex
      url: https://latex.codecogs.com/svg.image?
    html_preview: TRUE
    keep_html: TRUE
bibliography: RWM_Endnote_Library.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
---

@@

# Background and Objectives
This document details the modeling workflow implemented for estimating dissolved and equilibrium N2O gas concentrations and saturation ratios for all freshwater lakes and reservoirs in the lower 48 US states larger than 4 ha. Data for these estimates came from the 2017 Nation Lakes Assessment (NLA) survey, wherein waterbodies were sampled according to a spatially balanced, stratified, and unequal probability design. Stratification was among categories of lake size (surface area in hectares), WSA9 ecoregion, and US state (excluding AK and HI); and larger lakes were intentionally over-sampled relative to smaller lakes. In order for this sample to be useful for making inferences relevant to the population of interest, any estimates derived from it needed to be adjusted for potential biases arising from these design factors. It can be useful to consider the complete data likelihood when imagining how such biases may arise in sampling designs[@Link_Barker_2010; @Zachmann_etal_2022]. For the 2017 NLA survey, the complete data was considered to be all US lakes and reservoirs in the lower 48 states larger than 4 hectares. The survey samples, by comparison, are then considered a subset of the complete data with a known pattern of "missingess" assigned by the survey. For example, some lakes in the target population were missing from the sample not at random, but conditional on the design parameters. Their "missingness" is not random and, therefore, not ignorable when making inferences from the sample to the population of interest [@Gelman_etal_2014, Ch. 8; @Zachmann_etal_2022].

One model-based strategy to account for this selection bias is to include the survey design variables as predictors in a regression model and then adjust the resulting estimates based on a known target population distribution; a post-fitting adjustment referred to as poststratification [@Gelman_etal_2020, Ch. 17]. A particularly popular version of this approach uses multilevel regression models and is aptly referred to as multilevel regression and poststratification, or MRP [@Park_etal_2004; @Gelman_Little_1997; @Gelman_etal_2020, Ch. 17; @Kennedy_Gelman_2021]. Multilevel models are often recommended because they can provide regularized estimates along the design groupings, which can improve out-of-sample inferences [e.g., @Gelman_Little_1997; @Kennedy_Gelman_2021]. Estimates for group levels that may be missing from the sample, but are part of the population of interest, are also straightforward using the multilevel approach [@Gelman_etal_2020 Ch. 17; @McElreath_2020].

With a stratified, unequal probability design, such as the one used for the 2017 NLA, the selection design prescribes a number of random draws for each of a number of pre-specified groupings (i.e., sampling unit types). Within a particular stratification level or grouping, the draws may be considered a simple random sample (SRS) from that group type in the target population. A regression model conditioning the sample observations on the design grouping structure could, therefore, be parameterized to provide group-specific estimates that are unbiased with respect to the target population. To make inferences to the broader target population, however, those regression estimates would need to be re-weighted by the known proportions of units across groups in the target population. If the estimates are not adjusted in this manner, the estimates may be biased, particularly if the response of interest varies importantly across the grouping structure. For example, for the 2017 NLA survey, the proportion of samples prescribed to small lakes was much lower, relative to their proportion in the target population, compared to larger lakes, which were sampled in greater proportion relative to the target population. Small lakes are so numerous in the US that, if this unequal weighting by lake size weren't prescribed, nearly all the samples from the survey would be from small lakes and inferences regarding larger lakes could be highly uncertain for any reasonably achievable number of samples.  On the other hand, if lake size had no real influence on the N2O responses, the stratification across size class would make little difference with regard to inferences. Sampling any sizes of lakes would result in the same marginal estimates and standard errors.

For a recent applied example and helpful tutorial employing these concepts in a model-based approach using MRP, see @Kennedy_Gelman_2021. For an example of a recent application in the context of national surveys of environmental resources, see @Zachmann_etal_2022. In this study, a similar model-based approach was used to make population-level estimates from the 2017 NLA dissolved gas data. The specific workflow, data, models, and code used to make these estimates are documented in the remainder of this document. First, Bayesian multilevel regression models were used to fit the sample data to the design variable structure. Next, because eventual flux estimates needed to be estimated from lake-level measures (i.e., surface area), instead of predicting to a typical postratification table, predictions were made to each individual lake in the population of interest. This meant predicting to 465,897 individual natural and man made US lakes larger than 4 hectares in the lower 48 states. These predictions were assumed relevant to average conditions during the biological index period for each lake in 2017. The specific objective of the modeling effort was to provide population estimates for (1) dissolved and equilibrium N2O concentrations; (2) the N2O saturation ratio (i.e., dissolved N2O/equilibrium N2O); and (3) the proportion of under-saturated water bodies (i.e., saturation ratio < 1). As previously indicated, the gas estimates would also be used to estimate the total flux of N2O gas attributable to the target population of lakes over the index period. Saturation ratio was calculated as the ratio of dissolved to equilibrium N2O. Because dissolved and equilibrium N2O were observed on the same sample units (lake sites), models were fit to their joint distribution and the ratio estimates were assembled as a derived quantity. The response variable was, therefore, multivariate in order to account for any potential statistical dependencies between dissolved and equilibrium N2O due to, for example, common dependencies on geography. Although predictions of the mean marginal probabilities from separate models may have provided comparable estimates, a joint model allowing correlated errors was expected to better capture uncertainty and potentially improve out-of-sample predictions, should the variables be conditionally correlated [@Warton_etal_2015; @Poggiato_etal_2021]. All of the models fit were constructed using the `brms` package [@Burkner_2017] in `R` [@R_Core_Team_2021] as an interface to Stan, a software package for fitting fully Bayesian models via Hamiltonian Monte Carlo [HMC, @Stan_Development_Team_2018_a; @Stan_Development_Team_2018_b; @Stan_Development_Team_2018_c]. More specific details on the model structures is provided throughout the "Model fitting" section below.

# Data
As explained in a previous data munging document document
(https://github.com/USEPA/DissolvedGasNla/blob/master/scripts/dgIndicatorAnalysis.html), duplicate dissolved gas samples were collected at a depth of ~0.1m at designated index sites distributed across 1091 lakes nationwide, of which 95 were sampled twice as repeat visits. This subset of revisit sites was used as a test set for assessing model fit and out-of-sample performance.

Gas samples were analyzed via gas chromotography and concentrations were recorded to the nearest 0.001 nmol/L. As part of the survey design, each gas observation was indexed to an individual lake selected with unequal probability from 5 different lake size categories, $j \in j=1,...,J = 5$, according to surface area (ha), and from within a state, $k \in k=1,...,K = 48$, situated within an aggregated, WSA9 or Omernik ecoregion, $l \in l=1,...,L = 9$. All 9 WSA9 ecoregions were represented in the sample, including Xeric (XER), Western Mountain (WMT),  Northern Plains (NPL), Southern Plains (SPL), Temperate Plains (TPL), Coastal Plains (CPL), Upper Midwest (UMW), Northern Appalachian (NAP), and Southern Appalachian (SAP) regions. As shown below, the data from the initial and revisit samples were separately compiled into data frame objects in $\textbf{R}$, with $n=984$ and $n=95$ rows, respectively, of gas observations indexed to the survey design variables and several potentially relevant covariates.

## Import
The gas data and covariates were previously described and munged at 
https://github.com/USEPA/DissolvedGasNla/blob/master/scripts/dataMunge.html. That dataset was imported below.

```{r import_data, eval=FALSE, include=TRUE}
load( file = paste0( localPath,
              "/Environmental Protection Agency (EPA)/",
              "ORD NLA17 Dissolved Gas - Documents/",
              "inputData/dg.2021-02-01.RData")
      )

@@

A new data frame for modeling was constructed from the original file including only the variables of interest: (1) the N2O gas observations; (2) the survey design variables indexed to those observations; and (3) additional covariates considered potentially useful for improving the fit of the model. The modeling data frame below excluded the second-visit observations, which would later be used for model checking. Some variables from the imported data were renamed for convenience. In addition, the NO3 covariate was rounded according to the documented measurement precision and an alternative version was also created by log-transforming and re-coding the variable as an ordered factor with five levels. The left-most cut point separated observations below the detection limit from the completely observed samples. The remaining cut points were drawn at approximately equal distances in the positive direction along the log scale. Finally, one lake had no N2O gas information and was removed from the data frame.

```{r model_data, echo=TRUE, paged.print=TRUE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/dg.rda")

@@

A second data frame including only the second visit observations was constructed below. These data were later used as a "test set" to assess the out-of-sample performance of the model developed on the first-visit or "training set".

```{r test_data, echo=TRUE, paged.print=TRUE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/dg.rda")

@@

## Target population
Below. the NLA sampling frame was imported and filtered to include all lakes the target population. The resulting target population above included a total of 465,897 waterbodies.

```{r import_sample_frame, echo=TRUE}
df_pop <- read.csv(file = paste0(localPath,
              "/Environmental Protection Agency (EPA)/",
              "ORD NLA17 Dissolved Gas - Documents/",
              "inputData/NLA_Sample_Frame.csv"), header = T)

@@

Cross tabulations below describe the structure of the target population with respect to the survey design variables. The cross-tabulation shows that each ecoregion does not contain each state. Therefore, in the statistical sense, states were nested in ecoregions.

```{r frame_dimensions_1, echo=TRUE, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sframe.rda")

@@

The cross-tablulation below indicates that lake size category was nested in state (which was nested in ecoregion). That is, not every ecoregion:state contained every size category.

```{r frame_dimensions_4, echo=TRUE, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sframe.rda")

@@

Below, the sampling frame was munged to create a post-stratification table. There were 536 "types" or groupings of lakes in the population of interest with respect to the sampling design. The total counts of those lake types (n_lakes) and their proportions (prop_cell) relative to the counts in the target population were tabulated.

```{r filter_frame, echo=TRUE, message=FALSE, cache=TRUE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/sframe.rda")

@@

## Sample vs. population
Below, the lake distributions in the population of interest were compared to the proportions in the observed sample. There were 352 lake types in the sample compared to the 536 in the population of of interest. There were 984 observations distributed across these 352 lake types in the sample; and the number of samples was not distributed evenly across the types. Some cells were represented by as few as 1 lake. In total, 536-352 = 184 lake types in the population of interest were not represented in the sample.

```{r sample_cell_counts, echo=TRUE, message=FALSE, cache=TRUE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

Below, a graphical comparison was constructed to depict the distribution of cells in the population of interest _vs._ those in the sample.

```{r compare_sample_pop_cells, echo=FALSE, fig.align='center', fig.height=4, fig.width=10}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/pframe.rda")
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/samp_props.rda")

@@

Another comparison between population and sample was constructed by ecoregion. The samples were not balanced across ecoregions. Lakes in the Coastal Plains (CPL) ecoregion, for example, were clearly undersampled relative to their proportion of the population.

```{r eco_props_pop, eval=FALSE, include=TRUE, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/pframe.rda")

@@

A similar comparison by state was constructed below.

```{r state_props_pop, eval=FALSE, include=TRUE, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/pframe.rda")

@@

Finally, a comparison by lake size category is shown below. Small lakes were under-sampled relative to larger lakes.

```{r size_props_pop, eval=FALSE, include=TRUE, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/pframe.rda")

@@

## Sample-based estimates
Below are _naive_ estimates, based only on the sample, for national means for dissolved and equilibrium N2O and the saturation ratio.

```{r sample_summary_n2o}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

```{r sample_summary_n2oeq}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

```{r sample_summary_n2osat}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

Roughly 67% of lakes in the sample were under-saturated (i.e., saturation ratio < 1):

```{r sample_summary_propsat}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

Using only the sample observations again, a plot was constructed of the overall mean (dashed line) along with the ecoregion-specific means (black circles). The shaded areas indicate +/- 1 standard deviation. Neither dissolved N2O nor the saturation ratio were clearly structured by ecoregion, but there did appear to be some structure along this variable in the equilibrium N2O observations.

```{r sample_summary_eco, echo=FALSE, fig.align='center', fig.height=6, fig.width=8, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

The same summary by state is below.

```{r sample_summary_state, echo=FALSE, fig.align='center', fig.height=6, fig.width=8, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

Finally, the same summary by size category.

```{r sample_summary_size, echo=FALSE, fig.align='center', fig.height=6, fig.width=8, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

## Sample data exploration
Below, the distribution N2O concentrations for the sample was summarized using a density and rug plot. Note the natural log scale of the x-axis. Both the dissolved and equilibrium N2O data had considerable right skew even after the log transformation. This was not unexpected and has been noted in other studies [@Webb_etal_2019]. The saturation ratio was also skewed since it was derived from the other two observed variables (i.e., sat_ratio = n2o / n2o_eq).

```{r summary_N2O, echo=FALSE, fig.align='center', fig.height=8, fig.width=5, message=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

Below are plots of dissolved N2O vs. NO3. The first plot shows log(N2O) vs. log(NO3), as well as the ordinal categories assigned to NO3 (vertical lines). The leftmost vertical line is dashed and separates the NO3 observations below the detection limit. The trend is increasing and nonlinear on the log scale, with increasing variance in N2O as NO3 increased.

```{r summary_N2O_vs_NO3, echo=FALSE, fig.align='center', fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

Below are plots of dissolved N2O vs. NO3 for 6 quantiles of the surface temperature measurements (quantiles increasing from 1 to 6). This plot suggested that the NO3 effect on N2O may have been stronger in lakes with higher observed temperatures.

```{r summary_N2O_vs_NO3_surftemp, echo=FALSE, fig.align='center', fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

The next plot below shows the relationship between dissolved N2O and NO3 at 6 different quantiles (increasing 1 to 6) of the log-scaled lake surface area estimates.

```{r summary_N2O_vs_NO3_logarea, echo=FALSE, fig.align='center', fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
load("C:/Users/rmartin/OneDrive - Environmental Protection Agency (EPA)/Documents/AE_Reservoirs/DissolvedGasNla/modelFiles/df_model.rda")

@@

wherein $\varrho$ captured the group-level residual correlation between effects.
